---
title: "Elasticity Analysis on Avocado Supply and Demand"
author: "Hanchun Jiang"
date: "2021/11/10"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
```

```{r}
options(scipen=0, digits=7)
```

```{r}
avocado <- read_csv("https://raw.github.com/carolinejiang757/Elasticity-Analysis-of-Avocado-Supply-and-Demand/main/avocado.csv")
```
*The codes can be found at: https://github.com/carolinejiang757/Elasticity-Analysis-of-Avocado-Supply-and-Demand

## 1. Data Cleaning

**Do any data cleaning you think appropriate on the data set and report
any changes to the data that you make and why you made those changes.**

There are two main aspects of the raw data set that can be improved:
region and type. The original "region" column contains regions of
different levels, and the "type" column prevents me from analyzing the
two types together. Therefore, I made the following changes and made new
data sets:

**avocado:**

The original data set. A column "year" is added to store the year value
of a particular data entry.

**US:**

As the region values in the "region" column contain regions of different
level, I made the "US" data set specific to data with region value
"TotalUS".

**wei_avocado:**

As there is a type column in the original data frame, I made a weighted
full data frame where "volume" is the total volume of different types,
and "price" is the weighted price over two types.

**wei_US:**

As the region values in the "region" column contain regions of different
level, I broke "wei_avocado" into smaller data frames according to
region level. "wei_US" is the data frame for weighted national data.

**wei_big_region:**

This data set contains weighted data of big regions (defined as regions
larger than a state).

**wei_small_region:**

This data set contains weighted data of small regions (defined as
regions smaller than or equal to a state).

```{r}
avocado <-
  avocado %>% 
  rename(
    date = Date,
    avg_price = AveragePrice,
    total_volume = `Total Volume`,
    total_bags = `Total Bags`,
    small_bags = `Small Bags`,
    large_bags = `Large Bags`,
    xl_bags = `XLarge Bags`
    ) %>% 
  mutate(date = ymd(date), year = year(date))
```

```{r}
US <- filter(avocado, region == "TotalUS")
```

```{r include=FALSE}
unique(avocado$region)
```

```{r message=FALSE}
wei_avocado <- 
  avocado %>% group_by(region, date) %>%
      summarise(
                volume = sum(total_volume), 
                price = weighted.mean(avg_price, total_volume),
                year = year(date)
                )
```

```{r}
wei_avocado <- wei_avocado[-seq(1, nrow(wei_avocado), 2),]
```

```{r}
wei_US <- filter(wei_avocado, region == "TotalUS")
wei_big_region <- filter(wei_avocado, region %in% c("West", "Midsouth", "Northeast", "SouthCentral", "SouthEast", "GreatLakes", "NorthernNewEngland", "Plains"))
wei_small_region <- filter(wei_avocado, !region %in% c("West", "Midsouth", "Northeast", "SouthCentral", "SouthEast", "GreatLakes", "NorthernNewEngland", "Plains", "TotalUS"))
```

## 

## 2. Three Figures

**Create three figures that collectively do the best job of describing
and summarizing the data, i.e., after seeing your three figures, I
should say, "ok, now i understand the basic patterns in the avocado
market."**

### 1) Volume, Region and Time

This figure shows how the total volume of avocado consumed in the US
fluctuates over time. It also shows the composition of the total volume
by regions, given a certain time.

```{r}
ggplot(wei_big_region, aes(fill=region, y=volume, x=date, width = 7)) + 
    geom_bar(position="stack", stat="identity") +
   scale_y_continuous(
    name = "Total Volume US", labels = label_number(suffix = " M", scale = 1e-6)) +
  ggtitle("US Avocado Total Volume over Time, by Regions")
```

### 

### 2) Price, Type, and Time

This figure shows how the price of conventional and organic avocado in
the US fluctuates over time. It also shows the comparison of
conventional and organic avocado prices, given a certain time.

```{r message=FALSE}
colors <- c("Conventional" = "black", "Organic" = "red")
ggplot(US, aes(x = date, y = avg_price)) +
  geom_line(data = filter(US, type == "conventional"), aes(color = "Conventional")) +
  geom_line(data = filter(US, type == "organic"), aes(color = "Organic")) +
  ggtitle("US Avocado Price over Time, by Types") +
  labs(y = "Average Price",colour = "") +
  scale_color_manual(values = colors)
```

### 

### 3) Volume, Type and Time

This figure shows how the volume of conventional and organic avocado in
the US fluctuates over the four years.

\*Please note the conventional type corresponds to the left Y axis, and
the organic type corresponds to the right Y axis.

```{r}
coef1 <- 20
US02 <- mutate(US, total_volume = case_when(type == "organic" ~ total_volume * coef1, TRUE ~ total_volume))
ggplot(US02,
       aes(x = factor(year), y = total_volume, fill = type)) + 
  geom_boxplot() +
  scale_y_continuous(
    name = "Conventional Volume", labels = label_number(suffix = " M", scale = 1e-6),
    sec.axis = sec_axis(~./coef1, name = "Organic Volume", labels = label_number(suffix = " M", scale = 1e-6))
  ) +
  ggtitle("US Avocado Volume over Years, by Types")
```

## 

## 3. Run Regressions

**Run 1) time series, 2) cross-section, and 3) panel data regressions
with the log quantity of avocados as the left hand side variable and the
log price of avocados as the right hand side variable. Don't include any
extra control variables, other than year dummies and geographic dummies
if appropriate for the particular regression you are running.  Discuss
the coefficients you obtain. How would you describe what those
coefficients are in words?  How do they relate to economic parameters
people care about? Would you say the time series variation and the
cross-sectional variation are more driven by fluctuations in supply
factors or demand factors?**

### 1) Time Series

Time series data frame is the "wei_US" data frame created in Q1. It
shows how price and volume of avocado change over time in the US.

Time series data frame:

```{r}
wei_US
```

Time series regression:

```{r}
summary(
  lm(data = wei_US
     , log(volume) ~ log(price) + factor(year)
  )
)
```

### 

### 2) Cross Sectional

Cross sectional data frame is created as the follows. It compresses the
time dimension using average values, and shows how price and volume of
avocado differ in different small regions.

Cross sectional data frame:

```{r}
cross <- wei_small_region %>%
      group_by(region) %>%
      summarise(volume = mean(volume), price = mean(price) )
cross
```

Cross sectional regression:

```{r}
summary(
  lm(data = cross, log(volume) ~ log(price))
)
```

### 

### 3) Panel Data

Panel data frame is the "wei_small_region" data frame created in Q1. It
shows how price and volume of avocado change with two variables: time
and region.

Panel data frame:

```{r}
wei_small_region
```

Panel data regression:

```{r}
summary(
  lm(data = wei_small_region, log(volume) ~ log(price) + year + region
  )
)
```

### 

### 4) Interpretation

Coefficient for log(price) means how much log(volume) changes when
log(price) changes in one unit. This shows elasticity of the volume of
avocado over price change, though it's yet not clear whether this is
supply or demand elasticity.

Coefficient for the years means how log(volume) differs when in
different years. This shows how volumes of avocados are inherently
different between different years, due to varied demand and supply over
the years.

Coefficient for the regions means how log(volume) differs when in
different regions. This shows how volumes of avocados are inherently
different between different regions, due to varied demand and supply
between different regions.

### 

I think time series variation is more driven by supply. Looking at the
graph in Q4, which illustrates time series variation of volume and
price, I find that most of the times volume and price are moving in
opposite directions, indicating the impact of supply (more supply, lower
price; less supply, higher price).

Cross sectional variation, on the other hand, is more driven by demand.
Intuitively, besides regions where avocados are harvested, the amount of
avocodos transported and sold in each city is mainly determined by the
prediction of the demand for avocado in that city.

## 

## 4&6. Elasticity of Supply and Graph

**The department of agriculture wants your best estimate of the
elasticity of supply. Run whatever regression you think best gets at
that.**

**Explain why you made these choices. What estimate would you give for
the elasticity of supply? How confident are you in your estimate, i.e.
what standard error band would you put around your estimate? Make a
graph that best explains your finding.**

In the following graph, I found an area where volume and price are both
decreasing (red area). As volume and price are moving in the same
directions, this indicates a demand shock. Intuitively, this period
(2016-12-4 to 2016-12-18) is the period when people are preparing for
the Christmas, and as avocado is not among the traditional ingredients
for Christmas, and people tend to buy the its substitute goods, the
demand for avocado decreased, which caused a demand shock.

```{r message=FALSE}
Sys.setlocale("LC_TIME", "English")
```

```{r}
coef <- 10.5^7

ggplot() +
  geom_line(data = filter(wei_US, date >= as.Date("2016-6-1") & date <= as.Date("2017-6-1")), 
            aes(y = volume,x = date)) +
  geom_line(data = filter(wei_US, date >= as.Date("2016-6-1") & date <= as.Date("2017-6-1")), 
            aes(y = price * coef, x = date)) + 
  scale_y_continuous(
    name = "Volume", labels = label_number(suffix = " M", scale = 1e-6),
    sec.axis = sec_axis(~./coef, name = "Price")
  ) +
  geom_rect(aes(xmin = as.Date("2016-12-4"), xmax = as.Date("2016-12-18"), ymin=-Inf, ymax=Inf), fill = "red", alpha = 0.15) +
  geom_rect(aes(xmin = as.Date("2016-11-15"), xmax = as.Date("2017-1-15"), ymin=-Inf, ymax=Inf), fill = "blue", alpha = 0.15) +
  ggtitle("US Avocado Volume and Price over Time")
```

By running regression over the two months surrounding this demand shock
(blue area), and making a dummy variable which is "0" when not during
the shock and "1" when during the shock, I found out the change of
volume due to the shock by looking at the coefficient for the dummy. The
change of volume is -0.13540. Please note that this coefficient is not
very significant with a p value of 0.1, because this is a slight demand
shock, which caused a slight change in price and lasted for a short
period of time. However, this demand shock is already the most
significant one compared to other periods of time, as it is the longest
time span when volume and price are moving in the same direction.

```{r}
supply_elas <- filter(wei_US, date >= as.Date("2016-11-15") & date <= as.Date("2017-1-15"))

supply_elas <- mutate(supply_elas, shock = case_when(
  date %in% c(as.Date("2016-12-04"), as.Date("2016-12-11"), as.Date("2016-12-18")) ~ "1",
  TRUE ~ "0")
) 
```

```{r}
summary(
  lm(data = supply_elas, 
     log(volume) ~ log(price) + shock
  ))
```

I also need to find out the change of price caused by the shock. The
average price not during the shock minus the average price during the
shock is 0.075. Deviding the change of volume by the change of price, I
get the supply elasticity, which is 0.557.

```{r}
 (mean(filter(supply_elas, shock == "0")$price) - mean(filter(supply_elas, shock == "1")$price))/0.13540
```

The standard error given by the regression is 0.07. The actual standard
error should be higher due to many factors in reality not being ideal
(such as measurement error), and also due to the fact that the
coefficient is divided by another estimate (i.e. estimate for the change
in price). I will report a 0.2 standard error.

## 

## 5&6. Elasticity of Demand and Graph

**The department of agriculture wants your best estimate of the
elasticity of demand. Run whatever regression you think best gets at
that.**

**Explain why you made these choices. What estimate would you give for
the elasticity of demand? How confident are you in your estimate, i.e.
what standard error band would you put around your estimate? Make a
graph that best explains your finding.**

In the following graph, I marked a drastic volume decrease and
corresponding price increase with the red line and the shaded area. As
volume and price are moving in the opposite directions, this indicates a
supply shock.

The news below also confirms that the marked area indicates a supply
shock in around November, 2016 caused by a strike in Mexico and a
drought in California.

[\<https://www.forbes.com/sites/geoffwilliams/2016/10/31/how-the-avocado-shortage-is-affecting-chipotle-grocers-and-you/?sh=7344b52b5475>](https://www.forbes.com/sites/geoffwilliams/2016/10/31/how-the-avocado-shortage-is-affecting-chipotle-grocers-and-you/?sh=7344b52b5475){.uri}

```{r}
coef <- 10.5^7

ggplot() +
  geom_line(data = wei_US, aes(y = volume,x = date)) +
  geom_line(data = wei_US, aes(y = price * coef, x = date)) + 
  scale_y_continuous(
    name = "Volume", labels = label_number(suffix = " M", scale = 1e-6),
    sec.axis = sec_axis(~./coef, name = "Price")
  ) +
  geom_vline(xintercept = as.Date("2016-11-1"), color = "red", linetype = "dashed") +
  geom_rect(aes(xmin = as.Date("2016-9-1"), xmax = as.Date("2016-11-10"), ymin=-Inf, ymax=Inf), fill = "red", alpha = 0.15) +
  ggtitle("US Avocado Volume and Price over Time")
```

By running regression on this supply shock period, I found the demand
elasticity over price is -1.60877.

```{r}
demand_elas <- filter(wei_US, as.Date("2016-9-1") <= date & date <= as.Date("2016-11-10"))
summary(
  lm(data = demand_elas, log(volume) ~ log(price)
  )
)
```

The standard error given by the regression is 0.12635. The actual
standard error should be higher due to many factors in reality not being
ideal, such as measurement error. I will report a 0.2 standard error.

## 

### 7. Advice on Data Collection

As the most difficult and time consuming part during the data analysis
process is distinguishing the impact of supply from the impact of
demand, the Department of Agriculture should perform data collection in
ways that distinguish supply from demand.

Advice: For total US data, the Department of Agriculture should collect
data on how much avocados are produced, imported and exported in the US.
For regional data, they should collect data on how much avocados are
produced, transported into the region, and exported from the region. In
this way, the supply of avocado could be calculated, and it will be
easier to perform analysis related to supply and demand.
